Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-22 paper.tex
Up correciones N y efectos principales + render

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study examines the interplay between perceptions of economic
inequality and meritocracy in shaping preferences for market justice in
Chile from 2016 to 2023. Economic inequality has intensified globally,
raising concerns about its social impacts and the legitimacy of
market-based distributions of resources. By utilizing longitudinal data
from the Chilean Longitudinal Social Survey (ELSOC), the research
investigates how subjective assessments of inequality influence
attitudes towards social justice, focusing on the extent to which
individuals believe access to essential services like healthcare,
education, and pensions should depend on personal merit. The findings
suggest that as perceptions of economic inequality rise, individuals
tend to favor redistributive policies, although those with strong
meritocratic beliefs are more likely to justify existing inequalities.
The analysis also highlights the effect of major social movements in
Chile during this period, which have shifted public discourse and
potentially altered perceptions of inequality and meritocracy. This
paper contributes to the understanding of how individual beliefs about
fairness and equity evolve within a socio-political context
characterized by high inequality and extensive free-market principles.
\newline \textbf{Keywords}: Economic inequality, meritocracy, market
justice, Chile, public preferences
\end{abstract}

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit and effort. It is possible to trace
the concept to the understanding of Lane
(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
system of ``earned deserts,'' whereby individuals are seen as deserving
of a determined distributive outcome due to their effort and skills. In
contrast, political justice, emphasizes principles of equality and need,
which are often represented by the welfare state action through social
policies. An important remark is that the principles of market justice
rely on the assumption that markets are neutral and self-regulating
arenas, where individuals are treated fairly because they face the same
formal rules of engagement (\citeproc{ref-lane_market_1986}{Lane,
1986}). Consequently, the legitimacy of market justice stems from the
assumption that inequalities are not only inevitable but fair---so long
as the rules are transparent and opportunities are open. In this way,
market justice provides a moral justification for inequality by framing
it as a necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
foundations that explain public support for economic inequality. Over
time, this approach has been extended beyond income to include other
market-mediated outcomes, such as education, healthcare, and/or
pensions. For example, Von Dem Knesebeck et al.
(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Similarly, comparative studies have also
relied on these indicators to study the support for market-based
distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions.

@@

\(H1\): Higher perceived economic inequality is associated with higher
market justice preferences.

@@

This study draws on data from the Chilean Longitudinal Social Survey
(ELSOC), a panel study collected annually from 2016 to 2023. The survey
evaluates how individuals think, feel, and behave regarding conflict and
social cohesion in Chile. ELSOC employs a probabilistic, stratified,
clustered, multistage sampling design encompassing major urban centers
(Santiago, Valparaíso, and Concepción) and smaller cities. The target
population includes women and men aged 18--75 who are habitual residents
of private dwellings. The first wave included 2,927 participants,
representing northern and southern regions, covering 77\% of the
country's total population and 93\% of its urban population, with a
response rate of 62.4\% (\citeproc{ref-elsoc_estudio_2022}{ELSOC,
2022}).

The survey has been conducted yearly since 2016, except in 2020, when it
was suspended due to the pandemic. Waves 2016, 2017, 2018, 2019, 2022,
and 2023 used computer-assisted personal interviewing (CAPI), while a
reduced wave in 2021 employed computer-assisted telephone interviews
(CATI). Wave 3 included a refreshment sample (N = 1,519) to counter
attrition, but these cases are not used here to capture longer response
trends. Between wave 1 and wave 7, attrition amounted to 40\%, achieving
a final sample of N = 1,741. Longitudinal weights adjust for both the
sampling design and potential biases arising from systematic
non-participation. Further details on sampling, attrition, and weighting
can be found at \url{https://coes.cl/encuesta-panel/}, and the dataset
is publicly available at
\url{https://dataverse.harvard.edu/dataverse/elsoc}.

@@

\textbf{Market justice preferences}: The dependent variable in this
study is preferences for market justice. This construct is
operationalized through three items that capture how strongly
individuals justify conditioning access to social services---healthcare,
pensions, and education---on income. Specifically, the justification of
inequality in healthcare is assessed by the question: ``Is it fair in
Chile that people with higher incomes can access better healthcare than
people with lower incomes?'' The same question is posed for pensions and
education. In all cases, respondents indicate their level of agreement
on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
(``strongly agree''). Additionally, we include a composite measure of
``market justice preferences'', calculated as the average of these three
items (\(\alpha\) = 0.84). This index ranges from 1 to 5, with higher
values indicating stronger preferences for market justice (see
Table~\ref{tbl-summary1}).

@@

\textbf{Perception of economic inequality}: The main independent
variable refers to the perception of economic inequality, measured
through the perceived wage gap.
(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
from the salary gap between the perceived salaries of jobs at opposite
ends of the occupational hierarchy. Specifically, it relies on the
division between the perceived salary of a large-company president and
that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
2011}). Higher values of this term indicate a greater perception of
economic inequality between occupations located at the extremes of the
status continuum. This measure includes a logarithmic term in order to
adjust income magnitudes (usually fewer cases with high income):

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3774}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3113}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2075}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
\caption{Independent variables for the first wave
(2016)}\label{tbl-summary2}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Inequality gap perception & \begin{minipage}[t]{\linewidth}\raggedright
Mean (sd) : 3.7 (1.1)\\
min \textless{} med \textless{} max:\\
0.4 \textless{} 3.7 \textless{} 6.9\\
IQR (CV) : 1.6 (0.3)\strut
\end{minipage} & 296 distinct values &
\begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
People are rewarded for their efforts &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
169 (11.3\%)\\
693 (46.2\%)\\
263 (17.5\%)\\
328 (21.9\%)\\
48 ( 3.2\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
People are rewarded for their intelligence &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
134 ( 8.9\%)\\
617 (41.1\%)\\
294 (19.6\%)\\
401 (26.7\%)\\
55 ( 3.7\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
\end{longtable}

@@

Sociodemographic and attitudinal variables are included to control for
potential composition effects in the population. In terms of
sociodemographic characteristics, we incorporate per capita household
income quintile, educational level (1=Less than Universitary,
2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
been previously shown to influence market justice preferences
significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
attitudinal variables, we include political identification (1=Left,
2=Center, 3=Right, 4=No identification) and subjective social status
(ranging from 1 to 10) as they may confound the relationship between
market justice preferences and perceptions of inequality and meritocracy
(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
preferences for healthcare, pensions, and education from 2016 to 2023.
Each year presents stacked percentage frequencies, and the flows between
them reflect opinion changes among the same individuals from one year to
the next, given that we are using panel data. For instance, of the
40.8\% who strongly disagreed with justifying inequality in healthcare
in 2019, around 24.3\% maintained that position in 2022, while the
remaining 16.5\% shifted toward other response categories---primarily
moving into disagreement rather than strong disagreement. Overall, more
than half of the respondents exhibit a high level of disagreement
(disagree + strongly disagree) with inequality in these three social
service areas over time. Despite this general pattern, recent waves show
a slight decrease in disagreement and a corresponding rise in support
for inequality. Specifically, in healthcare and education, although
disagreement remains substantial, agreement (agree + strongly agree)
increased from 7.4\% and 7.2\% in 2019 to 13.1\% and 14.2\% in 2023,
respectively. This shift is most evident in pensions, where the combined
agree/strongly agree category grew by about 10 percentage points, from
16.9\% in 2016 to 28\% in 2023.

@@

Regarding the main variables of this study, Figure~\ref{fig-meanchange}
depicts their average changes over the years. We observe an increase in
the average level of market justice preferences in the most recent
waves, which begins in 2019. The highest average consistently appears
for perceived economic inequality, although this variable shows a
downward trend of roughly one point over time. Interestingly, while
perceptions of economic inequality declined in the latest measurements
(2022--2023), market justice preferences increased. The meritocracy
measures remain stable, though the perception that individuals are
rewarded for intelligence is slightly higher than the perception that
they are rewarded for their effort.

@@

\begin{center}
\scalebox{0.79}{
\begin{tabular}{l c c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
\hline
Intercept                            & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
                                     & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
Wave (Ref.= 2016)                    &                &                &                &                &                &                &                \\
                                     &                &                &                &                &                &                &                \\
\quad Wave 2017                      & $-0.167^{***}$ &                &                &                &                &                &                \\
                                     & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2018                      & $-0.025$       &                &                &                &                &                &                \\
                                     & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2019                      & $-0.046$       &                &                &                &                &                &                \\
                                     & $(0.027)$      &                &                &                &                &                &                \\
\quad Wave 2022                      & $0.285^{***}$  &                &                &                &                &                &                \\
                                     & $(0.028)$      &                &                &                &                &                &                \\
\quad Wave 2023                      & $0.284^{***}$  &                &                &                &                &                &                \\
                                     & $(0.027)$      &                &                &                &                &                &                \\
Wave                                 &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
                                     &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
Wave$^2$                             &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
                                     &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
Perception inequality (WE)           &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
                                     &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
Merit: Effort (WE)                   &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Merit: Talent (WE)                   &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
                                     &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
Merit: Effort (BE)                   &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
Merit: Talent (BE)                   &                &                &                &                &                & $0.040$        & $0.019$        \\
                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
\hline
Controls                             & No             & No             & No             & No             & No             & No             & Yes            \\
BIC                                  & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
Num. obs.                            & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
Num. groups: idencuesta              & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
Var: idencuesta (Intercept)          & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
Var: Residual                        & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
Var: idencuesta ola\_num             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
Cov: idencuesta (Intercept) ola\_num & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
\hline
\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
\end{tabular}
\caption{}
\label{table:coefficients}
\end{center}

>>>>

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic beliefs, and preferences for market justice in
Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with greater support for redistributive policies; however,
individuals with strong meritocratic convictions are more likely to
legitimize existing disparities. The study also considers the influence
of major social movements during this period, which appear to have
reshaped public discourse and perceptions of fairness. These findings
contribute to a deeper understanding of how beliefs about justice and
equity evolve in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts,'' whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility.

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
foundations that explain public support for economic inequality. Over
time, this approach has been extended beyond income to include other
market-mediated outcomes, such as education, healthcare, and/or
pensions. For example, Von Dem Knesebeck et al.
(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Similarly, comparative studies have also
considered these indicators to study the support for market-based
distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature {[}campbell\_policy\_2012;
pierson\_when\_1993{]}.

@@

\(H1\): Higher perceived economic inequality is associated with less
market justice preferences.

@@

This study draws on data from the Chilean Longitudinal Social Survey
(ELSOC), a panel study conducted annually from 2016 to 2023. The survey
assesses how individuals think, feel, and behave regarding social
conflict and cohesion in Chile. ELSOC employs a probabilistic,
stratified, clustered, multistage sampling design covering major urban
centers (Santiago, Valparaíso, and Concepción) as well as smaller
cities. The target population includes men and women aged 18 to 75 who
are habitual residents of private dwellings.

The survey has been conducted every year since 2016, except in 2020,
when it was suspended due to the COVID-19 pandemic. The first wave
included 2,927 participants from both northern and southern regions,
covering 77\% of the national population and 93\% of the urban
population, with a response rate of 62.4\%
(\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}). This study uses six
waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
excluded because a reduced version of the questionnaire omitted key
variables of interest. Between waves 1 and 6, panel attrition reached
40\%, resulting in a final two-level sample comprising N = 8,643
observations nested within N = 1,687 individuals. Longitudinal weights
are applied to adjust for both the sampling design and potential biases
from systematic non-response. Further details on sampling, attrition,
and weighting procedures are available at
https://coes.cl/encuesta-panel/, and the dataset is publicly accessible
at https://dataverse.harvard.edu/dataverse/elsoc.

@@

\textbf{Market justice preferences}: The dependent variable in this
study is preferences for market justice. This construct is
operationalized through three items that capture how strongly
individuals justify conditioning access to social services---healthcare,
pensions, and education--- basen on individual income. Specifically, the
justification of inequality in healthcare is assessed by the question:
``Is it fair in Chile that people with higher incomes can access better
healthcare than people with lower incomes?'' The same question is posed
for pensions and education. In all cases, respondents indicate their
level of agreement on a five-point Likert scale ranging from 1
(``strongly disagree'') to 5 (``strongly agree''). Additionally, we
include a composite measure of ``market justice preferences'',
calculated as the average of these three items (\(\alpha\) = 0.84). This
index ranges from 1 to 5, with higher values indicating stronger
preferences for market justice (see Table~\ref{tbl-summary1}).

@@

\textbf{Perception of economic inequality}: The main independent
variable refers to the perception of economic inequality, measured
through the perceived wage gap
(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
from the salary gap between the perceived salaries of jobs at opposite
ends of the occupational hierarchy. Specifically, it relies on the
division between the perceived salary of a large-company president and
that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
2011}). Higher values of this term indicate a greater perception of
economic inequality between occupations located at the extremes of the
status continuum. This measure includes a logarithmic term in order to
adjust income magnitudes (usually fewer cases with high income):

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3774}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3113}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2075}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
\caption{Independent variables ELSOC survey (descriptives for first wave
2016)}\label{tbl-summary2}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Label
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Stats / Values
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Freqs (\% of Valid)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Valid
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Inequality gap perception & \begin{minipage}[t]{\linewidth}\raggedright
Mean (sd) : 3.7 (1.1)\\
min \textless{} med \textless{} max:\\
0.4 \textless{} 3.7 \textless{} 6.9\\
IQR (CV) : 1.6 (0.3)\strut
\end{minipage} & 296 distinct values &
\begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
People are rewarded for their efforts &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
169 (11.3\%)\\
693 (46.2\%)\\
263 (17.5\%)\\
328 (21.9\%)\\
48 ( 3.2\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
People are rewarded for their intelligence &
\begin{minipage}[t]{\linewidth}\raggedright
1. Strongly desagree\\
2. Desagree\\
3. Neither agree nor desagre\\
4. Agree\\
5. Strongly agree\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
134 ( 8.9\%)\\
617 (41.1\%)\\
294 (19.6\%)\\
401 (26.7\%)\\
55 ( 3.7\%)\strut
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
1501\\
(100.0\%)\strut
\end{minipage} \\
\end{longtable}

@@

Sociodemographic and attitudinal variables are included to control for
potential composition effects in the population. In terms of
sociodemographic characteristics, we incorporate per capita household
income quintile, educational level (1=Less than Universitary,
2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
been previously shown to influence market justice preferences
significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
attitudinal variables, we include political identification (1=Left,
2=Center, 3=Right, 4=No identification) and subjective social status
(ranging from 1 to 10) as they may affect the relationship between
market justice preferences, perceptions of inequality, and meritocracy
(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
preferences for healthcare, pensions, and education from 2016 to 2023.
Each year presents stacked percentage frequencies, and the flows between
them reflect opinion changes among the same individuals from one year to
the next, given that we are using panel data. For instance, of the
40.8\% who strongly disagreed with justifying inequality in healthcare
in 2019, around 24.3\% maintained that position in 2022, while the
remaining 16.5\% shifted toward other response categories---primarily
moving into disagreement rather than strong disagreement. Overall, more
than half of the respondents exhibit a high level of disagreement
(disagree + strongly disagree) with inequality in these three social
service areas over time. Despite this general pattern, recent waves show
a slight decrease in disagreement and a corresponding rise in support
for market-justice inequality. Specifically, in healthcare and
education, although disagreement remains substantial, agreement (agree +
strongly agree) increased from 7.4\% and 7.2\% in 2019 to 13.1\% and
14.2\% in 2023, respectively. This shift is most evident in pensions,
where the combined agree/strongly agree category grew by about 10
percentage points, from 16.9\% in 2016 to 28\% in 2023.

@@

Regarding the main dependent and independent variables of this study,
Figure~\ref{fig-meanchange} depicts their average changes over the
years. We observe an increase in the average level of market justice
preferences in the most recent waves, which begins in 2019. The highest
average consistently appears for perceived economic inequality, although
this variable shows a downward trend of roughly one point over time.
Interestingly, while perceptions of economic inequality declined in the
latest measurements (2022--2023), market justice preferences increased.
The meritocracy measures remain stable, though the perception that
individuals are rewarded for intelligence is slightly higher than the
perception that they are rewarded for their effort.

@@

~

Model 1

Model 2

Model 3

Model 4

Model 5

Model 6

Model 7

Intercept

1.968***

2.015***

2.030***

2.028***

2.169***

1.456***

1.547***

~

(0.021)

(0.035)

(0.035)

(0.035)

(0.070)

(0.097)

(0.115)

Wave (Ref.= 2016)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Wave 2017

-0.167***

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2018

-0.025

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2019

-0.046

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~~~~~Wave 2022

0.285***

~

~

~

~

~

~

~

(0.028)

~

~

~

~

~

~

~~~~~Wave 2023

0.284***

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

Wave

~

-0.121***

-0.125***

-0.128***

-0.128***

-0.128***

-0.129***

~

~

(0.022)

(0.022)

(0.022)

(0.022)

(0.022)

(0.022)

Wave\^{}2

~

0.028***

0.028***

0.029***

0.029***

0.029***

0.029***

~

~

(0.003)

(0.003)

(0.003)

(0.003)

(0.003)

(0.003)

Perception inequality (WE)

~

~

-0.039***

-0.034***

-0.034***

-0.034***

-0.034***

~

~

~

(0.010)

(0.010)

(0.010)

(0.010)

(0.010)

Merit: Effort (WE)

~

~

~

0.070***

0.070***

0.069***

0.069***

~

~

~

~

(0.012)

(0.012)

(0.012)

(0.012)

Merit: Talent (WE)

~

~

~

0.027*

0.027*

0.027*

0.027*

~

~

~

~

(0.012)

(0.012)

(0.012)

(0.012)

Perception inequality (BE)

~

~

~

~

-0.042*

-0.006

-0.033

~

~

~

~

~

(0.018)

(0.018)

(0.018)

Merit: Effort (BE)

~

~

~

~

~

0.186***

0.178***

~

~

~

~

~

~

(0.032)

(0.031)

Merit: Talent (BE)

~

~

~

~

~

0.040

0.019

~

~

~

~

~

~

(0.032)

(0.031)

Controls

No

No

No

No

No

No

Yes

BIC

20780.594

20780.617

20781.113

20726.753

20736.627

20655.172

20749.047

Numb. obs.

8643

8643

8643

8643

8643

8643

8643

Num. groups: individuals

1687

1687

1687

1687

1687

1687

1687

Var: individuals (Intercept)

0.172

0.223

0.219

0.209

0.207

0.182

0.171

Var: Residual

0.530

0.499

0.498

0.495

0.495

0.495

0.495

Var: individuals, wave

~

0.011

0.010

0.009

0.009

0.009

0.009

Cov: individuals (Intercept), wave

~

-0.025

-0.024

-0.021

-0.020

-0.019

-0.019

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05.
